\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{booktabs}
\usepackage{hyperref}

\title{Learning Predictive World Models from Videos: A Survey on Generative Simulation and Physical Dynamics}
\author{TODO}
\date{TODO}

\begin{document}

\maketitle

\begin{abstract}
Predictive world models learned from videos aim to model how visual environments evolve and how future states depend on actions, latent controls, physical interactions, or evaluation criteria. Although recent progress in video generation has improved visual realism, visually plausible clips do not by themselves establish predictive usefulness, physical consistency, or controllability. This survey reviews learning predictive world models from videos through a concept-driven taxonomy rather than a chronological list of architectures. We organize the literature by modeling goal, representation, dynamics, conditioning signal, and evaluation protocol. Under this view, the field spans action-conditioned and stochastic video prediction, object-centric and relational physical dynamics, latent world models for planning and policy learning, diffusion-based image-space simulators, interactive generative environments, and benchmarks for physical consistency and control utility. The survey emphasizes the recurring tension between compact abstractions that support efficient decision making and image-space models that may preserve control-relevant visual detail. It also highlights evaluation gaps: frame reconstruction, video realism, and task return measure different properties and should not be treated as interchangeable evidence of world-model validity. We conclude by identifying open challenges in long-horizon prediction, physical-law generalization, controllability, object and causal abstraction, and evaluation beyond surface realism.
\end{abstract}

\input{sections/01_introduction}
\input{sections/02_background}
\input{sections/03_taxonomy}
\input{sections/04_video_prediction}
\input{sections/05_object_centric_physical_dynamics}
\input{sections/06_latent_world_models}
\input{sections/07_diffusion_interactive_world_models}
\input{sections/08_evaluation_open_challenges}
\input{sections/09_conclusion}

\bibliographystyle{plain}
\bibliography{../bib/references}

\end{document}
