\section{Introduction}
\label{sec:introduction}

Predictive world models learned from video aim to model how visual environments evolve, how they respond to interventions, and how their future states can be used for reasoning or control. This goal is increasingly important because video provides dense evidence about motion, contact, occlusion, object persistence, and agent-environment interaction, while requiring less manual state specification than classical simulators. Early action-conditioned video prediction already framed visual forecasting as a transition problem in which future observations depend on candidate actions, and robot interaction prediction extended this idea to physical manipulation by predicting how actions transform observed pixels \cite{oh2015_action_conditional_video_prediction_prov,finn2016_physical_interaction_video_prediction_prov}. In this sense, video prediction is not only an image-generation problem. It is also a route toward learning predictive models of environments from observation.

The field covered by this survey is broader than generic video generation because a world model is judged by the role its predictions play, not only by whether its samples look plausible. A video generator may produce coherent clips, yet a predictive world model should support at least one stronger use case: forecasting under actions, representing uncertainty over possible futures, exposing object and relational structure, enabling planning or policy learning, generating controllable environments, or evaluating physical consistency. Stochastic video prediction illustrates the need to model multiple plausible futures rather than a single deterministic rollout \cite{babaeizadeh2018_sv2p_prov}. Latent world models make the distinction sharper by compressing visual observations into predictive states for planning, latent imagination, or policy evolution rather than treating future-frame fidelity as the final objective \cite{ha2018_world_models_prov,hafner2019_planet_prov,hafner2020_dreamer_prov}.

The central difficulty is that visual realism, predictive usefulness, physical consistency, and interactivity are related but non-equivalent objectives. Pixel-level fidelity can miss whether predictions preserve information needed for action; object-oriented prediction can support planning in structured tasks even when the evaluation target is not ordinary video quality; and an image-space diffusion world model can be evaluated through the behavior of an agent trained inside the learned simulator \cite{oh2015_action_conditional_video_prediction_prov,janner2019_o2p2_prov,alonso2024_diamond_prov}. Conversely, visually plausible generated videos can still violate physical commonsense, so realism alone should not be treated as evidence of a physically valid simulator \cite{bansal2025_videophy_prov}. A recurring theme of this survey is therefore the tension between abstraction and fidelity: compact latent states can make planning efficient, while image-space generative models may preserve visual details that matter for control \cite{hafner2019_planet_prov,hafner2020_dreamer_prov,alonso2024_diamond_prov}.

This survey focuses on learned predictive models from videos that are relevant to generative simulation and physical dynamics. The scope includes action-conditioned video prediction, stochastic future modeling, object-centric and relational dynamics, latent world models for planning and control, diffusion-based image-space world models, interactive generative environments, and evaluation protocols for physical consistency and controllability. It excludes pure image generation, language-only world modeling, classical physics simulators without learned visual prediction, and robotics work that does not include a video or world-modeling component. Within the current evidence base, some modern diffusion-video, benchmark, and latent-control papers are indexed but not yet supported by completed notes; this draft therefore treats them as coverage targets rather than established evidence.

We organize the literature with a concept-driven taxonomy rather than a chronological paper list. The taxonomy uses five axes: modeling goal, representation, dynamics, conditioning, and evaluation. Modeling goals distinguish frame prediction, stochastic forecasting, physical rollout, planning, policy learning, interactive generation, and physical evaluation. Representations range from pixels and motion fields to stochastic latents, object graphs, recurrent latent states, diffusion/noise spaces, and video tokens. Dynamics mechanisms include action-conditioned recurrent predictors, motion-transform predictors, variational video models, graph interaction networks, recurrent state-space models, conditional diffusion rollouts, and latent-action dynamics \cite{oh2015_action_conditional_video_prediction_prov,finn2016_physical_interaction_video_prediction_prov,babaeizadeh2018_sv2p_prov,battaglia2016_interaction_networks_prov,watters2017_visual_interaction_networks_prov,hafner2019_planet_prov,hafner2020_dreamer_prov,alonso2024_diamond_prov,bruce2024_genie_prov}. Conditioning separates passive extrapolation from prediction under actions, robot commands, policies, text prompts, or learned latent controls, while evaluation asks whether the model is tested for visual fidelity, planning utility, controllability, physical commonsense, or long-horizon behavior.

This survey makes four contributions. First, it clarifies the boundary between video generation and predictive world modeling by emphasizing prediction under state history, action, interaction, and evaluation criteria. Second, it synthesizes the field through a taxonomy that compares assumptions, representations, dynamics models, conditioning signals, and evaluation protocols across pixel-level, object-centric, latent, diffusion-based, and interactive approaches. Third, it highlights the abstraction-fidelity trade-off that runs through the literature, connecting compact latent planning models with image-space generative simulators and structured physical dynamics \cite{battaglia2016_interaction_networks_prov,watters2017_visual_interaction_networks_prov,janner2019_o2p2_prov,hafner2019_planet_prov,hafner2020_dreamer_prov,alonso2024_diamond_prov}. Fourth, it identifies evaluation gaps, especially the need to test world models beyond frame metrics and task return using physical consistency, causal or counterfactual reasoning, controllability, and long-horizon robustness \cite{oh2015_action_conditional_video_prediction_prov,alonso2024_diamond_prov,bruce2024_genie_prov,bansal2025_videophy_prov}.

The remainder of the survey follows this taxonomy. The background section defines the scope and distinguishes video prediction from world modeling. Section~\ref{sec:taxonomy} introduces the taxonomy axes and conceptual families. Section~\ref{sec:video_prediction} reviews video prediction and visual dynamics, including action-conditioned and stochastic formulations. Section~\ref{sec:object_centric_physical_dynamics} discusses object-centric and relational physical dynamics. Section~\ref{sec:latent_world_models} covers latent world models for planning and control. Section~\ref{sec:diffusion_interactive_world_models} examines diffusion-based and interactive world models. Section~\ref{sec:evaluation_open_challenges} discusses evaluation protocols and open challenges, including physical consistency, controllability, long-horizon prediction, and the limits of visual realism.

% Concise Claim-Evidence Audit:
% - Major claims:
%   Predictive world models from video are broader than generic video generation; action conditioning, stochasticity, object structure, latent planning, diffusion rollout, and interactivity define different world-model roles; visual realism, predictive usefulness, physical consistency, and interactivity are distinct evaluation targets; the survey contributes a concept-driven taxonomy and an evaluation-focused synthesis.
% - Supporting papers:
%   oh2015_action_conditional_video_prediction_prov; finn2016_physical_interaction_video_prediction_prov; babaeizadeh2018_sv2p_prov; ha2018_world_models_prov; hafner2019_planet_prov; hafner2020_dreamer_prov; battaglia2016_interaction_networks_prov; watters2017_visual_interaction_networks_prov; janner2019_o2p2_prov; alonso2024_diamond_prov; bruce2024_genie_prov; bansal2025_videophy_prov.
% - Weak or missing evidence:
%   Detailed claims about modern diffusion-video models, broader physical-reasoning benchmarks, newer latent-control variants, and newer interactive diffusion world models remain deferred until the corresponding notes are completed.
% - Revision notes:
%   Revisit the introduction after completing notes for P010, P011, P014--P016, P020--P022, and P033--P040; add or update Figure 1 to match the final taxonomy; strengthen benchmark claims once the evaluation notes are evidence-ready.
