\section{A Concept-Driven Taxonomy}
\label{sec:taxonomy}

This survey organizes predictive world models by the role their predictions play, rather than by a chronological list of architectures. The key distinction is that generic video generation can be judged mainly by visual plausibility or prompt adherence, whereas a predictive world model should support some form of environment prediction under state history, actions, latent controls, planning objectives, or physical-evaluation criteria. Early action-conditioned video prediction already made future frames depend on agent actions, latent world models used compact predictive states for planning or policy learning, object-centric models exposed entities and relations, DIAMOND used conditional diffusion as an image-space environment model, Genie learned controllable latent actions from unlabeled video, and VideoPhy showed why visually plausible generated videos still require separate physical-consistency evaluation \cite{oh2015_action_conditional_video_prediction_prov,hafner2019_planet_prov,hafner2020_dreamer_prov,battaglia2016_interaction_networks_prov,watters2017_visual_interaction_networks_prov,janner2019_o2p2_prov,alonso2024_diamond_prov,bruce2024_genie_prov,bansal2025_videophy_prov}. This motivates a taxonomy based on modeling goal, representation, dynamics, conditioning, and evaluation.

\subsection{Taxonomy Axes}

\paragraph{Modeling goal.}
The first axis asks what the learned model is meant to make possible: frame prediction, stochastic forecasting, physical rollout, planning, policy learning, interactive generation, or evaluation. This axis separates models that mainly extrapolate videos from models whose predictions are used for intervention or decision making. For example, action-conditioned Atari prediction and robot pushing forecast future observations under actions, while PlaNet and Dreamer use learned dynamics to support planning or behavior learning \cite{oh2015_action_conditional_video_prediction_prov,finn2016_physical_interaction_video_prediction_prov,hafner2019_planet_prov,hafner2020_dreamer_prov}. The modeling goal therefore determines which failures matter: a visually plausible rollout may still be inadequate if it does not preserve control-relevant details or physical interactions.

\paragraph{Representation.}
The second axis asks what state space carries the prediction. Pixel-space and motion-transform models predict or transform image observations directly \cite{oh2015_action_conditional_video_prediction_prov,finn2016_physical_interaction_video_prediction_prov}. Stochastic video predictors add latent variables to represent multiple plausible futures \cite{babaeizadeh2018_sv2p_prov}. Latent world models compress images into compact recurrent states for planning or imagined policy learning \cite{ha2018_world_models_prov,hafner2019_planet_prov,hafner2020_dreamer_prov}. Object-centric models instead represent entities, relations, or object-factorized latent states \cite{battaglia2016_interaction_networks_prov,watters2017_visual_interaction_networks_prov,janner2019_o2p2_prov}. Image-space diffusion world models occupy a different point in this design space because they retain generated observations as the rollout medium, which DIAMOND studies in Atari \cite{alonso2024_diamond_prov}.

\paragraph{Dynamics.}
The third axis asks how the model rolls the world forward. The reviewed methods illustrate several distinct mechanisms: action-conditioned feature transformations, pixel-motion transformations, variational stochastic video prediction, recurrent latent-state transitions, graph interaction dynamics, object-oriented perception-physics-rendering pipelines, autoregressive token transitions, feature-space prediction, conditional diffusion dynamics, causal diffusion sequence modeling, and token dynamics with latent actions \cite{oh2015_action_conditional_video_prediction_prov,finn2016_physical_interaction_video_prediction_prov,babaeizadeh2018_sv2p_prov,hafner2019_planet_prov,hafner2020_dreamer_prov,battaglia2016_interaction_networks_prov,janner2019_o2p2_prov,wu2024ivideogpt,assran2025vjepa2_prov,alonso2024_diamond_prov,chen2024diffusion_forcing_prov,bruce2024_genie_prov}. This axis is not redundant with representation: two models may both operate on images while using different transition mechanisms, and two latent models may differ in whether they support online planning, actor-critic imagination, modular policy evolution, token prediction, or feature-space prediction.

\paragraph{Conditioning and interaction.}
The fourth axis asks what information controls the rollout. Passive predictors condition mainly on past frames, but world-model use cases often require future states to depend on interventions. Actions enter explicitly in Atari prediction, robot manipulation prediction, PlaNet, Dreamer, and DIAMOND \cite{oh2015_action_conditional_video_prediction_prov,finn2016_physical_interaction_video_prediction_prov,hafner2019_planet_prov,hafner2020_dreamer_prov,alonso2024_diamond_prov}. Genie broadens the axis by learning discrete latent actions from unlabeled video rather than using ground-truth action labels \cite{bruce2024_genie_prov}. Conditioning is therefore the main boundary between merely predicting what is likely to happen and asking what could happen under alternative controls.

\paragraph{Evaluation.}
The fifth axis asks what property of the model is actually tested. Pixel metrics such as reconstruction loss, PSNR, and SSIM can be informative for frame prediction, but they do not by themselves establish planning utility, controllability, or physical correctness. Action-conditioned prediction can be evaluated through control usefulness, object-oriented prediction can be evaluated through planning success, and DIAMOND evaluates an image-space diffusion world model through Atari control performance \cite{oh2015_action_conditional_video_prediction_prov,janner2019_o2p2_prov,alonso2024_diamond_prov}. Physical-consistency evaluation adds a further test: VideoPhy evaluates whether generated videos satisfy both caption semantics and physical commonsense, and organizes prompts around material-interaction categories such as solid-solid, solid-fluid, and fluid-fluid interactions \cite{bansal2025_videophy_prov}. For this survey, evaluation is therefore a cross-cutting axis rather than a separate paper family.

\subsection{Conceptual Families}

The axes above define non-exclusive conceptual families. A paper may belong to several families at once, so the taxonomy should be read as an analytic map rather than a set of disjoint bins.

\paragraph{Action-conditioned pixel and motion prediction.}
This family covers models that predict future visual observations in image space while conditioning on known actions or robot commands. It provides an early bridge from passive video prediction to world modeling because the predicted future changes with the intervention. Action-conditioned Atari prediction uses control inputs to forecast high-dimensional game observations, while robot interaction prediction uses pixel-motion transformations to model how actions move objects \cite{oh2015_action_conditional_video_prediction_prov,finn2016_physical_interaction_video_prediction_prov}. This family prepares the later video-prediction section by showing why action conditioning and downstream utility matter before the survey moves to more abstract world states.

\paragraph{Stochastic future modeling.}
This family addresses cases where a single deterministic future is an inadequate target. SV2P uses latent variables to generate different plausible futures and reports design trade-offs between time-varying and time-invariant stochastic latents \cite{babaeizadeh2018_sv2p_prov}. The discussion remains narrow because it is used here to identify multi-modality, latent-variable use, and evaluation with sampled futures as distinct concerns rather than to survey all stochastic video models.

\paragraph{Object-centric and relational physical dynamics.}
This family represents physical systems through entities, relations, or object-factorized latent states. Interaction Networks operate on explicit object-relation graphs, Visual Interaction Networks connect visual perception to relational dynamics, and O2P2 learns object-oriented perception, physics, and rendering for prediction and planning \cite{battaglia2016_interaction_networks_prov,watters2017_visual_interaction_networks_prov,janner2019_o2p2_prov}. The claims should remain qualified because much of the cited evidence comes from controlled or synthetic settings, and some models assume object states, segmentation, or supervised state targets. Even with these limits, the family prepares the structured-dynamics section by making object identity, relations, hidden causes, and planning relevance explicit.

\paragraph{Latent world models for planning and control.}
This family shifts the predictive state from pixels to compact latent dynamics. The recurrent world-model formulation factorizes perception, memory, and control into VAE, MDN-RNN, and controller modules; PlaNet plans in recurrent latent state space; and Dreamer trains behavior through imagined latent trajectories \cite{ha2018_world_models_prov,hafner2019_planet_prov,hafner2020_dreamer_prov}. iVideoGPT and V-JEPA 2 broaden the family toward token-based interactive world models and self-supervised feature-space prediction, respectively \cite{wu2024ivideogpt,assran2025vjepa2_prov}. This family prepares the planning-and-control section by separating online planning, amortized policy learning, policy optimization inside learned or hallucinated environments, token dynamics, and feature-space prediction. It also exposes an evaluation gap: strong control return does not automatically prove physical faithfulness in open-ended video domains.

\paragraph{Image-space and diffusion-based world models.}
This family asks when strong image-space generative models can function as predictive simulators. DIAMOND models environment dynamics with a conditional diffusion model over observations and actions, and evaluates the resulting world model through Atari agent performance \cite{alonso2024_diamond_prov}. Diffusion Forcing adds a causal-diffusion sequence-modeling point that connects next-token prediction with full-sequence diffusion \cite{chen2024diffusion_forcing_prov}. Broader claims about video diffusion as a general world-modeling paradigm should remain cautious unless they are tied to prediction, control, rollout, or physical evaluation. This family prepares the diffusion section by framing the trade-off between compact abstraction and preservation of control-relevant visual details.

\paragraph{Interactive and embodied generative environments.}
This family focuses on controllable video worlds. It includes models with explicit actions, such as action-conditioned Atari and robot-pushing predictors, and models with learned latent controls, such as Genie \cite{oh2015_action_conditional_video_prediction_prov,finn2016_physical_interaction_video_prediction_prov,bruce2024_genie_prov}. LAPA and iVideoGPT broaden the discussion toward latent-action pretraining and token-based interactive world modeling, while Cosmos marks the platform-scale world foundation model framing \cite{ye2025lapa_prov,wu2024ivideogpt,nvidia2025cosmos}. Genie shows that controllability can be induced from unlabeled video through latent actions, but its actions are not yet equivalent to semantically grounded real-world interventions. This family prepares the interactive-environments section by separating known actions, robot commands, learned latent actions, token controls, and downstream imitation or control evaluations.

\paragraph{Evaluation, physical consistency, and benchmarks.}
Evaluation is a cross-cutting family because every modeling branch can look different depending on what is measured. Frame metrics test visual prediction, planning success tests action usefulness, control return tests policy utility, and physical-consistency benchmarks test whether generated dynamics obey commonsense constraints \cite{oh2015_action_conditional_video_prediction_prov,janner2019_o2p2_prov,alonso2024_diamond_prov,bansal2025_videophy_prov}. VideoPhy is especially important for this survey boundary because it treats visual plausibility and physical correctness as separate evaluation targets \cite{bansal2025_videophy_prov}. Newer indexed evaluation papers extend this space toward action-centric physical commonsense, video-as-world-model judging, measurement-based physical evaluation, broader world-generation evaluation, and causal embodied scenarios \cite{bansal2025videophy2_prov,li2025worldmodelbench_prov,zhang2025morpheus_prov,duan2025worldscore,cai2026whatifworld_prov}. This family prepares the later evaluation section by making clear that world-model correctness cannot be reduced to generic video quality.

\begin{table*}[t]
\centering
\scriptsize
\setlength{\tabcolsep}{2pt}
\renewcommand{\arraystretch}{1.15}
\caption{Concept-driven taxonomy matrix used in this survey. Families are non-exclusive: a single paper may contribute to multiple rows.}
\label{tab:taxonomy_branches}
\begin{tabular}{p{0.095\textwidth} p{0.16\textwidth} p{0.12\textwidth} p{0.13\textwidth} p{0.115\textwidth} p{0.12\textwidth} p{0.12\textwidth} p{0.12\textwidth}}
\toprule
Family & Representative methods & Representation & Transition mechanism & Conditioning & Evaluation & Strengths & Limitations \\
\midrule
Action-conditioned pixel/motion prediction & Action-conditioned Atari prediction \cite{oh2015_action_conditional_video_prediction_prov}; robot interaction prediction \cite{finn2016_physical_interaction_video_prediction_prov} & Pixels, features, motion fields & Autoregressive prediction; learned pixel transformations & Past frames plus known actions & Frame metrics; control or task utility where available & Directly connects visual prediction to interventions & Short horizons; pixel metrics can miss action usefulness \\
Stochastic future modeling & SV2P \cite{babaeizadeh2018_sv2p_prov} & Pixels plus stochastic latents & Variational recurrent prediction & Past frames, optional actions, sampled latents & PSNR/SSIM; sampled-future evaluation & Represents multiple plausible futures & Uncertainty quality and decision utility remain hard to test \\
Object-centric physical dynamics & Interaction Networks \cite{battaglia2016_interaction_networks_prov}; Visual Interaction Networks \cite{watters2017_visual_interaction_networks_prov}; O2P2 \cite{janner2019_o2p2_prov} & Object states, graphs, object latents & Relational or object-oriented dynamics plus rendering/planning modules & Objects, relations, visual observations, task goals & Rollout error; planning success; structured generalization & Makes entities, relations, and hidden causes explicit & Often assumes controlled settings, object states, masks, or proposals \\
Latent planning and control & Recurrent World Models \cite{ha2018_world_models_prov}; PlaNet \cite{hafner2019_planet_prov}; Dreamer \cite{hafner2020_dreamer_prov}; iVideoGPT \cite{wu2024ivideogpt}; V-JEPA 2 \cite{assran2025vjepa2_prov} & Compact latent states, RSSM states, VAE/RNN latents, tokens, feature states & Recurrent state-space transitions; latent imagination; autoregressive token prediction; feature-space prediction & Observations, actions, rewards, policies & Return; planning success; behavior-oriented evaluation & Efficient decision making in learned predictive state & Return can miss physical validity; latents may discard control-relevant detail \\
Diffusion/interactive world models & DIAMOND \cite{alonso2024_diamond_prov}; Diffusion Forcing \cite{chen2024diffusion_forcing_prov}; Genie \cite{bruce2024_genie_prov}; LAPA \cite{ye2025lapa_prov}; Cosmos \cite{nvidia2025cosmos} & Images, diffusion/noise states, video tokens, latent actions & Conditional diffusion rollout; causal diffusion; token dynamics; latent-action dynamics & Observation-action history, sequence context, latent actions, robot commands, language & Control return; controllability; video quality; physical evaluation where available & Preserves visual detail and broadens controllable generation & Compute cost, ambiguous latent controls, and physical validity remain open \\
Evaluation and benchmarks & VideoPhy \cite{bansal2025_videophy_prov}; WorldModelBench \cite{li2025worldmodelbench_prov}; VideoPhy-2 \cite{bansal2025videophy2_prov}; Morpheus \cite{zhang2025morpheus_prov}; WorldScore \cite{duan2025worldscore}; What-If World \cite{cai2026whatifworld_prov} & Prompts, videos, scores, human judgments, physical measurements & Benchmark protocols and evaluator models rather than predictive dynamics & Tasks, prompts, labels, actions, changed conditions & Human judgment; automatic judging; physical measurements; causal/counterfactual tasks & Separates visual plausibility from world-model validity & Fragmented domains; automatic evaluators require validation \\
\bottomrule
\end{tabular}
\end{table*}

This taxonomy prepares the remaining survey sections by assigning each body section a question rather than a paper list. The video-prediction section asks when future-frame prediction becomes useful for world modeling; the stochastic section asks how models represent multiple futures; the structured-dynamics section asks when object and relation biases help physical prediction; the latent-control section asks how compact states support planning or policy learning; the diffusion section asks when image-space generation is worth its cost; the interactive-environments section asks how control enters video generation; and the evaluation section asks how to test physical consistency, controllability, and utility beyond visual realism.
