\section{Video Prediction and Visual Dynamics}
\label{sec:video_prediction}

Video prediction provides an early formulation of predictive visual world modeling, but only under a specific interpretation: the model is useful for this survey when it learns how visual observations evolve under history, actions, or uncertainty. In the evidence-ready papers reviewed in this section, the predicted state is primarily an image, feature map, or pixel-motion field, and evaluation often begins with reconstruction quality. This does not make video prediction equivalent to a full world model. Its relevance increases when predictions are conditioned on candidate actions, used to reason about possible interaction outcomes, or evaluated for utility beyond frame reconstruction. Action-conditioned Atari prediction illustrates this bridge by forecasting high-dimensional observations under discrete actions rather than only extrapolating a passive video stream \cite{oh2015_action_conditional_video_prediction_prov}.

\subsection{From Future-Frame Prediction to Visual Dynamics}

Future-frame prediction becomes world-model-relevant when it is treated as visual dynamics rather than only as video synthesis. The distinction is not that one line generates images and the other does not; both may produce future frames. The difference is the role of the generated frame: predictive visual dynamics asks whether the next observation follows from the previous observations and the conditioning variables defined by the task. In early action-conditioned work, future frames depend on both visual history and action input, so the model functions as a visually grounded transition predictor over agent observations \cite{oh2015_action_conditional_video_prediction_prov}. This framing prepares the later sections, which move the transition model from pixels into objects, relations, latent states, or image-space generative simulators.

Deterministic frame prediction exposes a first limitation of pixel-space visual dynamics: a single rollout can be useful, but it cannot represent multiple plausible futures. P004 and P005 provide evidence for the image-space and long-horizon limitations of deterministic action-conditioned prediction, including difficulty with small task-critical details and degradation over longer rollouts \cite{oh2015_action_conditional_video_prediction_prov,finn2016_physical_interaction_video_prediction_prov}. P006 motivates the related uncertainty problem more directly: when the future is multi-modal, a single deterministic target can average over alternatives and produce blurred predictions \cite{babaeizadeh2018_sv2p_prov}. Thus video prediction is an important starting point for world modeling, but its deterministic pixel-level form leaves open how to represent uncertainty, hidden causes, and longer-term consequences.

\subsection{Action-Conditioned Prediction and Physical Interaction}

Action conditioning is the key step that connects video prediction to control and interaction. A passive predictor estimates what is likely to happen next from visual history, whereas an action-conditioned predictor estimates what may happen under a specified action sequence. In Atari, this makes future observations depend on the discrete action sequence and allows predicted frames to be evaluated in relation to control usefulness, not only pixel error \cite{oh2015_action_conditional_video_prediction_prov}. The conceptual shift is that the predictor no longer models video as an autonomous stream. It models part of the agent-environment loop.

Robot video prediction extends this idea from game observations to physical interaction. In robot pushing, the model receives visual history and future robot actions, then predicts how the scene will change under those actions. Rather than reconstructing all future appearance from scratch, the approach in the completed notes predicts pixel motion through transformation mechanisms such as DNA, CDNA, and STP, which reuse observed pixels while learning how actions move them \cite{finn2016_physical_interaction_video_prediction_prov}. This design is intended to exploit the persistence of local appearance while modeling action-induced motion. At the same time, the resulting model remains an image-space predictor: its masks and transformations may reveal weak object-like structure, but they do not by themselves provide explicit object identities, hidden physical states, or a planning-ready causal model.

\subsection{Stochastic Futures and Long-Horizon Uncertainty}

Stochastic video prediction addresses one important limitation of deterministic visual prediction: future observations can be multi-modal. SV2P adds latent variables to an autoregressive video predictor so that different latent samples can generate different plausible futures from the same observed context \cite{babaeizadeh2018_sv2p_prov}. This changes the role of the predictor from producing one best future to sampling a family of possible rollouts. For world modeling, this matters because different plausible futures may correspond to different interaction outcomes, although the current evidence for downstream decision consequences remains limited.

The completed evidence also shows that stochasticity introduces modeling and optimization problems of its own. SV2P uses a variational objective and staged training because naive optimization can ignore the latent variables, and it studies time-varying and time-invariant latents as different ways to represent uncertainty over a rollout \cite{babaeizadeh2018_sv2p_prov}. This should be read as a paper-specific design observation rather than a general law: the current note base supports stochastic video prediction mainly through P006. The broader lesson is more cautious. Adding latent randomness is useful only if the model actually uses it, and evaluation must distinguish meaningful uncertainty from superficially different sampled frames.

\subsection{Evidence Gaps in Architecture-Oriented Video Prediction}

The completed evidence for this section currently covers action-conditioned Atari prediction, robot pixel-motion prediction, and SV2P-style stochastic prediction. The paper index also includes recurrent spatiotemporal predictors, simplified video-prediction baselines, motion/content factorization, physics-informed prediction, masked-token prediction, learned-prior stochastic prediction, and diffusion-based video prediction; however, the corresponding notes for P020 and P023--P028 are not yet evidence-ready. The draft therefore treats these methods as coverage targets and does not use them as support for detailed claims about baseline strength, planning-oriented prediction, physical-dynamics disentanglement, or diffusion-formulated video prediction.

Within the evidence-ready set, pixel-level prediction remains limited as a world-modeling objective. Pixel losses and frame metrics can reward short-horizon visual accuracy without testing whether the model preserves control-relevant details or remains useful over long rollouts. The completed notes support this concern from several angles: action-conditioned Atari prediction shows that frame quality alone can miss control usefulness, robot video prediction remains vulnerable to deterministic blur and long-horizon degradation, and stochastic video prediction relies partly on sampled-future evaluation \cite{oh2015_action_conditional_video_prediction_prov,finn2016_physical_interaction_video_prediction_prov,babaeizadeh2018_sv2p_prov}. The broader survey revisits this evaluation issue in settings where prediction is tested through planning or control, such as object-oriented prediction and image-space diffusion world modeling \cite{janner2019_o2p2_prov,alonso2024_diamond_prov}.

\subsection{Transition Toward Structured and Latent World Models}

The video-prediction line supplies both a foundation and a warning for predictive world models from video. Its foundation is the idea that a learned model can roll observations forward from visual history, actions, and sometimes stochastic latents. Its warning is that visually plausible prediction is not the same as a structured, controllable, or physically reliable simulator. The next sections therefore examine complementary responses along the taxonomy axes: object-centric models introduce entities and relations, latent world models learn compact predictive states for planning or policy learning, diffusion world models revisit image-space simulation, and interactive generative environments introduce explicit or learned controls \cite{janner2019_o2p2_prov,hafner2019_planet_prov,hafner2020_dreamer_prov,alonso2024_diamond_prov,bruce2024_genie_prov}. The role of video prediction in this survey is to establish the predictive problem that these later structured, latent, and generative simulators refine.

% Claim-Evidence Audit:
% - Claim: Video prediction is an early formulation of predictive visual world modeling when it rolls observations forward under visual history, actions, or uncertainty.
%   Supporting papers: P004; P005; P006
%   Citation keys: oh2015_action_conditional_video_prediction_prov; finn2016_physical_interaction_video_prediction_prov; babaeizadeh2018_sv2p_prov
%   Evidence source: C001; C002; C004; section packet 04; completed notes P004, P005, P006
%   Confidence: medium
%   Missing evidence: Later action-conditioned and architecture-oriented video-prediction notes would broaden the claim.
% - Claim: Action-conditioned video prediction connects passive future-frame prediction to interaction because predicted observations depend on candidate actions.
%   Supporting papers: P004
%   Citation keys: oh2015_action_conditional_video_prediction_prov
%   Evidence source: C001; C019; section packet 04; completed note P004
%   Confidence: high
%   Missing evidence: More completed notes on later action-conditioned predictors.
% - Claim: Robot video prediction connects visual prediction to physical interaction through action-conditioned pixel-motion transformations.
%   Supporting papers: P005
%   Citation keys: finn2016_physical_interaction_video_prediction_prov
%   Evidence source: C002; section packet 04; completed note P005
%   Confidence: high
%   Missing evidence: Broader motion/content and robot prediction notes are incomplete.
% - Claim: Deterministic pixel-level predictors have long-horizon and detail-preservation limits in the completed action-conditioned and robot-prediction evidence.
%   Supporting papers: P004; P005
%   Citation keys: oh2015_action_conditional_video_prediction_prov; finn2016_physical_interaction_video_prediction_prov
%   Evidence source: section packet 04; completed notes P004 and P005
%   Confidence: medium
%   Missing evidence: More completed notes on deterministic video prediction baselines are needed.
% - Claim: Stochastic latent-variable video prediction addresses multi-modal futures and introduces optimization choices such as staged training and latent temporal structure.
%   Supporting papers: P006
%   Citation keys: babaeizadeh2018_sv2p_prov
%   Evidence source: C004; C005; C006; section packet 04; completed note P006
%   Confidence: high for SV2P; medium as a broader field claim
%   Missing evidence: P028 and other stochastic-video notes are incomplete.
% - Claim: Pixel-level prediction and frame metrics are not sufficient to establish world-model usefulness.
%   Supporting papers: P004; P005; P006; P009; P012
%   Citation keys: oh2015_action_conditional_video_prediction_prov; finn2016_physical_interaction_video_prediction_prov; babaeizadeh2018_sv2p_prov; janner2019_o2p2_prov; alonso2024_diamond_prov
%   Evidence source: C003; section packet 04; claim evidence map
%   Confidence: high for the limitation; medium for broad evaluation implications
%   Missing evidence: Dedicated metric and benchmark notes remain incomplete.
% - Claim: Later object-centric, latent, diffusion, and interactive families can be organized as complementary responses to pixel-level video prediction limits.
%   Supporting papers: P009; P001; P002; P012; P013
%   Citation keys: janner2019_o2p2_prov; hafner2019_planet_prov; hafner2020_dreamer_prov; alonso2024_diamond_prov; bruce2024_genie_prov
%   Evidence source: taxonomy.md; C007; C011; C015; C019
%   Confidence: medium
%   Missing evidence: More completed notes are needed for modern diffusion-video and interactive world-model families.
