\section{Diffusion-Based and Interactive World Models}
\label{sec:diffusion_interactive_world_models}

Diffusion-based and interactive world models reopen the question of whether world simulation should preserve rich image-space detail or compress experience into compact latent states. The preceding section emphasized latent dynamics for planning and policy learning; this section considers a complementary direction in which high-capacity generative video models are used either as predictive simulators or as controllable visual environments. The shift is motivated by a simple tension: latent abstraction can make planning efficient, but aggressive compression may discard small visual details that matter for control, while image-space generation can preserve more appearance detail at higher computational and evaluation cost \cite{hafner2019_planet_prov,hafner2020_dreamer_prov,alonso2024_diamond_prov}. For this reason, diffusion models are relevant to world modeling, but they should not be equated with world models unless their generations are connected to prediction, action conditioning, rollout, controllability, or physically meaningful evaluation.

\subsection{Video Diffusion as Predictive Generation}

Video Diffusion Models and MCVD are relevant to the lineage behind image-space world models, with MCVD connecting diffusion to prediction, generation, and interpolation \cite{ho2022_video_diffusion_models_prov,voleti2022_mcvd_prov}. This section uses them only to mark the diffusion-video lineage. It does not make a fine-grained comparison of their architectures, datasets, empirical gains, video fidelity, temporal modeling, or prediction results.

The distinction between a generative video model and a world model is central to this survey. A visually plausible video generator may learn temporal correlations, object motion, and scene transitions, yet still fail to answer intervention questions such as what would happen under a different action, policy, or latent control. Earlier action-conditioned video prediction already made this boundary explicit by conditioning future frames on Atari actions, while robot interaction prediction made future observations depend on robot commands and physical contact \cite{oh2015_action_conditional_video_prediction_prov,finn2016_physical_interaction_video_prediction_prov}. Diffusion video generation becomes world-model relevant when it inherits this predictive or interactive role: the model must support rollouts under conditioning signals, not merely produce plausible videos from prompts or past frames.

\subsection{Conditional and Causal Diffusion Dynamics}

DIAMOND provides the strongest supported example here for diffusion as a world-modeling mechanism. Rather than treating diffusion only as an unconditional or text-conditioned video generator, DIAMOND models environment dynamics directly in image space, generating the next observation from a history of observations and actions \cite{alonso2024_diamond_prov}. Rollout is autoregressive: generated observations become part of the future context, and the learned simulator is used as the environment in which a reinforcement-learning agent is trained. This makes the diffusion model a transition model over visual observations, not only a video synthesis model.

This design changes the abstraction-fidelity trade-off established by latent world models. PlaNet and Dreamer demonstrate that compact recurrent latent states can support planning and imagined policy learning efficiently \cite{hafner2019_planet_prov,hafner2020_dreamer_prov}. DIAMOND instead argues that preserving image-space visual details can improve downstream reinforcement learning performance in Atari, suggesting that some information discarded by compact or discrete latent representations may be control-critical \cite{alonso2024_diamond_prov}. The claim should remain narrow: the cited evidence is Atari-centered, and it supports the importance of visual detail for some control settings rather than a general rule that image-space diffusion is always preferable.

Diffusion Forcing adds a second diffusion design point by connecting next-token prediction with full-sequence diffusion \cite{chen2024diffusion_forcing_prov}. In this survey it is used as evidence that the boundary between autoregressive sequence modeling and diffusion-style denoising is becoming a design axis for world models. The appropriate comparison is architectural: causal diffusion is complementary to DIAMOND-style conditional diffusion, not a replacement for full-sequence diffusion or proof of solved long-horizon physical simulation.

\subsection{Latent Actions and Interactive Control}

Interactive world models extend the diffusion and video-generation discussion from predicting likely futures to producing controllable futures. In action-labeled settings, this idea is visible in Atari video prediction, robot pushing, and DIAMOND: the future observation depends on a known intervention signal \cite{oh2015_action_conditional_video_prediction_prov,finn2016_physical_interaction_video_prediction_prov,alonso2024_diamond_prov}. Genie broadens the setting by learning controllability from unlabeled video through a spatiotemporal tokenizer, a latent action model, and a MaskGIT-style dynamics model \cite{bruce2024_genie_prov}. Instead of assuming ground-truth actions, it infers discrete latent actions that can be supplied at inference time to steer generated video.

Genie's contribution is conceptually important because it separates interaction from externally labeled control. Genie is trained from video-only data without ground-truth action labels, and its learned latent actions can support downstream imitation in a narrow CoinRun setting \cite{bruce2024_genie_prov}. At the same time, latent actions are not equivalent to physically grounded interventions. They may make a generated environment playable or steerable, but their semantics require post hoc interpretation, and the available evidence does not establish robust long-horizon physical consistency. Thus, Genie is best read as evidence that controllability can emerge from video-only learning, not as proof that unlabeled video alone yields physically reliable world simulators.

LAPA extends the latent-action theme into the vision-language-action setting and is used here as a reference point for latent actions, video pretraining, and embodied control \cite{ye2025lapa_prov}. The survey treats it as a design question rather than a settled empirical conclusion: latent action pretraining may reduce reliance on direct action annotation, but it does not eliminate the need to connect learned latent actions to robot actions or embodied policy learning. iVideoGPT is similarly useful as a token-based interactive world-model bridge between VideoGPT-style generation and controllable sequence modeling \cite{wu2024ivideogpt}.

\subsection{Large-Scale World Foundation Model Efforts}

Large-scale video-generation systems have also entered the world-model discussion. Cosmos is included here as an industrial-scale world foundation model platform for physical AI, not as evidence that large video models already solve physical simulation \cite{nvidia2025cosmos}. Its relevance is that it makes the world-model framing explicit at platform scale and shifts part of the discussion from single benchmark agents toward reusable generative simulation infrastructure.

This framing should remain cautious. A platform called a world foundation model still has to be evaluated for prediction under interventions, temporal consistency, controllability, and physical validity. Technical reports and platform papers can identify an important research direction, but they should not be treated as proof that video generation has become reliable physical simulation. For that reason, Cosmos is discussed alongside evaluation benchmarks rather than used as a stand-alone demonstration of physical correctness.

\subsection{Newer Interactive Diffusion Directions}

VideoWorld and Vid2World are newer papers at the boundary between video generation, diffusion, and interaction \cite{ren2025_videoworld_prov,huang2026_vid2world_prov}. They are relevant to the section goal, especially for the question of how passive video models become controllable environments. This section cites them only as examples of this emerging boundary and does not use them to support detailed claims about mechanisms, benchmarks, action spaces, or empirical results.

This evidence gap matters because interaction is not a single property. A model may be conditioned on real actions, robot commands, learned latent actions, text prompts, or policies, and each conditioning signal implies a different notion of controllability. Real actions can be tied to environment interventions, while latent actions may support steering without direct semantic grounding. Diffusion adapters, token dynamics, or video-only latent actions may bridge these regimes, but the survey should distinguish them carefully rather than treating all controllable generation as equivalent. The supported comparison is therefore conservative: action-conditioned prediction and DIAMOND support known-action rollout, while Genie supports latent-action interactive generation \cite{oh2015_action_conditional_video_prediction_prov,finn2016_physical_interaction_video_prediction_prov,alonso2024_diamond_prov,bruce2024_genie_prov}.

\subsection{Realism, Controllability, and Physical Validity}

The strongest limitation of diffusion and interactive video worlds is that visual realism, controllability, temporal consistency, and physical validity do not necessarily improve together. DIAMOND evaluates a diffusion world model through Atari control performance, but this does not establish whether image-space diffusion produces long-horizon causal or physical correctness in more realistic embodied domains \cite{alonso2024_diamond_prov}. Genie evaluates controllability and downstream imitation, but its latent actions are not semantically grounded real-world controls and physical faithfulness is not the main evaluation target \cite{bruce2024_genie_prov}. Cosmos, LAPA, Diffusion Forcing, iVideoGPT, VideoWorld, and Vid2World broaden the design space, but they should be treated cautiously when discussing physical-law generalization or robust embodied transfer \cite{nvidia2025cosmos,ye2025lapa_prov,chen2024diffusion_forcing_prov,wu2024ivideogpt,ren2025_videoworld_prov,huang2026_vid2world_prov}.

VideoPhy makes this evaluation problem explicit for generative video. It treats physical commonsense as a separate target from semantic or visual plausibility and reports that the best evaluated text-to-video model satisfies both caption and physical-law criteria on only a minority of instances \cite{bansal2025_videophy_prov}. This supports the survey's main caution: a model that generates convincing videos is not automatically a physically valid world model. For diffusion-based simulators, the corresponding open question is how to evaluate whether generated rollouts preserve causal structure, conservation-like behavior, object permanence, or counterfactual validity.

\subsection{Transition to Evaluation and Open Challenges}

Diffusion and interactive world models therefore sharpen the next section's evaluation agenda. They combine attractive properties from several earlier branches: image-space detail from video prediction, action dependence from embodied prediction, rollout from world-model RL, and controllability from interactive generation. Yet each property introduces a different failure mode. High visual fidelity can hide physical errors; action conditioning can be narrow or benchmark-specific; latent actions can be controllable but semantically ambiguous; and autoregressive or causal-diffusion rollouts can accumulate errors even when short clips look plausible.

The resulting open challenge is not to choose between latent abstraction and image-space generation once and for all, but to evaluate when each representation supports the intended world-model use. For control, the model should preserve details that affect decisions. For interaction, it should respond reliably to actions or latent controls. For physical simulation, it should maintain temporal and causal consistency beyond surface realism. For open-ended video generation, it should be judged by physical commonsense and not only by perceptual quality. These criteria motivate the next section, which treats evaluation as a multi-axis problem spanning fidelity, controllability, task utility, physical consistency, and long-horizon behavior \cite{oh2015_action_conditional_video_prediction_prov,alonso2024_diamond_prov,bruce2024_genie_prov,bansal2025_videophy_prov}.
