\section{Evaluation and Open Challenges}
\label{sec:evaluation_open_challenges}

Evaluation is the point at which video prediction, generative simulation, and world modeling must be separated. A video model can reconstruct frames, produce visually plausible samples, obtain high task return, or satisfy a prompt without necessarily learning a physically valid, controllable, or causally useful model of the world. The preceding sections therefore motivate a multi-axis evaluation view: reconstruction fidelity tests local visual prediction, generative video metrics test distributional visual quality, planning or control returns test task usefulness, physical benchmarks test whether dynamics obey commonsense or physical constraints, causal and counterfactual tasks test reasoning about interventions, and controllability tests whether a model responds reliably to actions or latent controls. No single axis is sufficient for the survey's target notion of a predictive world model.

\subsection{Visual and Reconstruction Metrics}

Reconstruction metrics remain useful, but they measure only a narrow part of world-model behavior. Early action-conditioned video prediction used pixel error and qualitative rollouts, yet frame prediction quality alone can underestimate usefulness for control \cite{oh2015_action_conditional_video_prediction_prov}. Robot video prediction similarly used PSNR and SSIM to compare future-frame quality, but its own limitations point to blur, short horizons, and the need to evaluate decision usefulness rather than only image similarity \cite{finn2016_physical_interaction_video_prediction_prov}. These examples show why reconstruction fidelity is best treated as a diagnostic for local observation prediction, not as evidence that a model has learned physical dynamics, controllable interventions, or task-relevant state.

Generative video quality metrics address a different question from world-model validity. Frechet Video Distance is a metric for evaluating generative video quality \cite{unterthiner2018_fvd_prov}, and several generative or interactive models report video-quality metrics in addition to task-specific measures. This section does not rely on detailed claims about how FVD is computed or how it correlates with human perception. More importantly, even a distributional video metric would not by itself test action sensitivity, causal validity, physical-law generalization, or long-horizon controllability. Visual quality and world-model correctness therefore need to be reported as related but distinct evaluation targets.

\subsection{Task Utility: Planning, Control, and Interaction}

Planning and control evaluations test whether a learned model supports decisions, not merely whether it predicts pixels. O2P2 provides a clear example: its object-oriented representation supports downstream planning better than an object-agnostic video predictor baseline in a tower-building task \cite{janner2019_o2p2_prov}. Latent world models evaluate learned dynamics through control returns, with PlaNet planning in latent space and Dreamer learning behavior through latent imagination \cite{hafner2019_planet_prov,hafner2020_dreamer_prov}. DIAMOND evaluates a diffusion world model through Atari agent performance, making control return an explicit test of an image-space learned simulator \cite{alonso2024_diamond_prov}. These evaluations are essential because they reveal whether predictions preserve information that matters for action.

Control utility is still not equivalent to general world validity. Reward-defined tasks can be narrow, benchmark-specific, and insensitive to physical errors outside the policy's visited region. A learned simulator may support a policy in Atari or a structured block-stacking task while remaining weak as an open-ended physical model. Interactive controllability adds another axis: Genie learns latent actions from unlabeled video and evaluates controllability and downstream imitation in a narrow setting, but its latent actions are not semantically grounded real-world controls \cite{bruce2024_genie_prov}. Thus, planning success, control return, imitation performance, and controllability should be interpreted as complementary tests rather than interchangeable evidence.

\subsection{Human-Judgment Physical Commonsense Evaluation}

Physical consistency evaluation asks whether generated or predicted videos obey the constraints implied by objects, materials, and interactions. VideoPhy evaluates generated videos for both semantic adherence and physical commonsense, organizes prompts around solid-solid, solid-fluid, and fluid-fluid interactions, and reports that the best evaluated text-to-video model satisfies caption and physical-law criteria on only a minority of instances \cite{bansal2025_videophy_prov}. This directly supports the survey's warning that visual realism should not be equated with world-model correctness. The benchmark is also useful conceptually because it treats physical commonsense as a measurable property rather than an informal qualitative impression.

VideoPhy-2 is included as a related action-centric physical-commonsense benchmark for video generation \cite{bansal2025videophy2_prov}. Its role here is to mark an expansion from material-interaction prompts toward action execution and action plausibility. This survey does not use VideoPhy-2 to make detailed claims about exact failure rates, evaluated models, or automatic-evaluator performance. The cautious synthesis is that human-judgment physical commonsense evaluation is becoming more fine-grained, while remaining sensitive to prompt design, annotation protocol, and benchmark scope.

\subsection{Automated Physics-Adherence Detection}

Automatic evaluators are promising but require careful validation. VideoPhy introduces VideoCon-Physics as an automatic evaluator and reports stronger agreement with human judgments than several compared multimodal baselines on its benchmark \cite{bansal2025_videophy_prov}. Physical commonsense judgments can remain partly qualitative and depend on prompt design and annotation protocol. For world models, this means automatic scores can help scale evaluation but should not replace human inspection, task rollouts, or intervention-based tests.

WorldModelBench strengthens this evaluation axis by explicitly judging video generation models as world models using criteria such as instruction following and physics adherence \cite{li2025worldmodelbench_prov}. This is a useful addition because it frames evaluation around whether generated videos satisfy world-model-like constraints rather than only video quality. WorldModelBench is best read as a benchmark direction for automated or semi-automated world-model assessment, not as evidence that automated physical evaluation is solved.

\subsection{Measurement-Based Physical Evaluation}

Measurement-based evaluation provides a complementary route to physical validity. Morpheus is included as a benchmark that connects video generative model evaluation to real physical experiments and conservation-law-guided metrics \cite{zhang2025morpheus_prov}. This style of evaluation is important because it attempts to compare generated dynamics with measured physical behavior rather than relying only on perceived plausibility or prompt adherence.

The contribution of measurement-based evaluation should also be framed cautiously. Real physical experiments can give stronger grounding for specific physical phenomena, but they may cover a restricted set of objects, motions, viewpoints, and measurement assumptions. Morpheus therefore complements human-judgment benchmarks such as VideoPhy and VideoPhy-2; it does not replace them. A robust evaluation suite should combine automatic metrics for scale, human judgment for semantic and commonsense validity, measurement-based tests where physical variables can be tracked, and task-specific control or planning evaluation where the model is used for action.

\subsection{Causal and Counterfactual Evaluation}

Causal and counterfactual evaluation is necessary because world models are often used to ask intervention questions. Object-centric and relational models provide partial evidence for this direction: Interaction Networks explicitly model object relations, Visual Interaction Networks can infer hidden physical causes such as invisible objects and unknown mass from visual effects, and O2P2 uses object-oriented prediction for planning \cite{battaglia2016_interaction_networks_prov,watters2017_visual_interaction_networks_prov,janner2019_o2p2_prov}. CLEVRER, PHYRE, and Physion are relevant benchmark families for causal video reasoning, physical reasoning, and physical prediction \cite{bakhtin2019_phyre_prov,yi2020_clevrer_prov,bear2021_physion_prov}. This section cites them as benchmark context but does not make detailed claims about their task formats, splits, metrics, or model results.

What-If World is included as an emerging causal benchmark for general world models in embodied scenarios \cite{cai2026whatifworld_prov}. Its inclusion marks the need to evaluate how models respond when physical details, actions, or embodied conditions are changed, rather than only asking whether a plausible forward video can be generated. This survey treats it as an emerging direction rather than a mature standard or a replacement for earlier causal and physical-reasoning benchmarks.

\subsection{Benchmark Coverage and Evidence Gaps}

The current benchmark coverage is richer but still fragmented. VideoPhy supports claims about physical commonsense evaluation for generated videos \cite{bansal2025_videophy_prov}. VideoPhy-2, Morpheus, WorldModelBench, and What-If World extend the evaluation map toward action-centric physical commonsense, measurement-based physical tests, world-model-oriented video judgment, and causal intervention evaluation \cite{bansal2025videophy2_prov,zhang2025morpheus_prov,li2025worldmodelbench_prov,cai2026whatifworld_prov}. WorldScore broadens the lens further by evaluating world generation beyond the specific physical-commonsense setting \cite{duan2025worldscore}. These citations establish benchmark coverage, not detailed comparative conclusions about model rankings or solved evaluation.

PHYRE, CLEVRER, Physion, FVD, PhyGenBench, How Far Is Video Generation from World Model, and Physics-IQ remain relevant to the broader evaluation landscape \cite{bakhtin2019_phyre_prov,yi2020_clevrer_prov,bear2021_physion_prov,unterthiner2018_fvd_prov,meng2025_phygenbench_prov,kang2025_physical_law_perspective_prov,motamed2025_physics_iq_prov}. Here, they are cited to indicate evaluation categories rather than to support detailed claims about what current world models prove on each benchmark. This distinction matters: citing a benchmark's existence is different from claiming what it proves about current world models.

\subsection{Open Challenges}

Long-horizon consistency remains a central challenge for predictive world models. Pixel-space predictors suffer from compounding rollout errors, latent models depend on the quality of imagined dynamics, object-centric models often operate in controlled settings, and diffusion rollouts raise open questions about causal and physical correctness beyond Atari \cite{oh2015_action_conditional_video_prediction_prov,hafner2020_dreamer_prov,watters2017_visual_interaction_networks_prov,alonso2024_diamond_prov}. The challenge is not only to make long videos look coherent, but to preserve state identity, contact outcomes, occluded variables, and action consequences over extended rollouts. Systematic long-horizon physical evaluation across realistic videos remains an open problem.

Physical-law generalization is still under-evaluated relative to visual realism. Structured models such as Interaction Networks and Visual Interaction Networks suggest that object and relation biases can support generalization and hidden-cause inference in controlled physical systems \cite{battaglia2016_interaction_networks_prov,watters2017_visual_interaction_networks_prov}. VideoPhy shows that generated videos can fail physical commonsense even when visually plausible \cite{bansal2025_videophy_prov}. Newer benchmarks such as PhyGenBench, How Far Is Video Generation from World Model, Physics-IQ, VideoPhy-2, WorldModelBench, and Morpheus are relevant to physical-law and physical-commonsense evaluation \cite{meng2025_phygenbench_prov,kang2025_physical_law_perspective_prov,motamed2025_physics_iq_prov,bansal2025videophy2_prov,li2025worldmodelbench_prov,zhang2025morpheus_prov}. This survey uses them as indicators of the evaluation direction rather than as evidence for detailed comparative conclusions.

Controllability and interactivity require evaluation protocols that go beyond passive generation. Action-conditioned Atari prediction, robot interaction prediction, DIAMOND, and Genie all make future observations depend on actions, robot commands, or latent actions \cite{oh2015_action_conditional_video_prediction_prov,finn2016_physical_interaction_video_prediction_prov,alonso2024_diamond_prov,bruce2024_genie_prov}. Yet these controls have different semantics: real actions correspond to environment interventions, robot commands are embodied and task-specific, and latent actions may be steerable without being physically interpretable. Future benchmarks should measure whether controls produce stable, repeatable, and semantically meaningful changes over time, rather than only whether generated clips vary.

Object and causal abstraction remains unresolved because useful world models must often infer hidden state from raw video. The reviewed object-relational models support object structure as a useful inductive bias, but also show important limitations: Interaction Networks assume explicit object graphs, Visual Interaction Networks rely on supervised state targets in synthetic environments, O2P2 uses segments or proposals, and Genie learns latent controls without explicit physical semantics \cite{battaglia2016_interaction_networks_prov,watters2017_visual_interaction_networks_prov,janner2019_o2p2_prov,bruce2024_genie_prov}. A central open problem is to learn objects, relations, hidden variables, and intervention semantics from weakly supervised or unlabeled video while preserving usefulness for planning and control.

Evaluation beyond FVD is necessary because video distribution quality does not test all world-model properties. A future evaluation suite should report reconstruction or prediction fidelity, distributional video quality, control return, controllability, physical consistency, causal and counterfactual reasoning, and long-horizon robustness. The exact metric set for such a suite is not established. The key principle is that frame metrics can miss control usefulness, task return can miss physical validity, and visually plausible generated videos can fail physical commonsense \cite{oh2015_action_conditional_video_prediction_prov,alonso2024_diamond_prov,bansal2025_videophy_prov}.

Scaling video models does not by itself guarantee physical grounding. The survey has traced a movement from pixel prediction to latent control, object-centric dynamics, diffusion simulators, and interactive video environments; each step increases capability along some axes while leaving other axes under-tested. DIAMOND suggests that preserving image-space details can matter for control, while PlaNet and Dreamer show why compact latent states are useful for efficient planning or imagination \cite{hafner2019_planet_prov,hafner2020_dreamer_prov,alonso2024_diamond_prov}. The open question is how to combine scale, abstraction, interaction, and physical grounding without letting visual realism substitute for evidence of causal world understanding.

Sim-to-real and embodied deployment are important stress tests for video world models. O2P2 includes evidence that an object-oriented learned representation can support planning and transfer to a real robotic setting in a structured task, while robot video prediction studies action-conditioned futures from real interactions \cite{janner2019_o2p2_prov,finn2016_physical_interaction_video_prediction_prov}. These examples are important but narrow. Broader embodied deployment would require models that handle contact, uncertainty, changing viewpoints, object identity, actuator limits, and distribution shift. The survey therefore treats reliable real-world simulation from video-trained world models as an open challenge rather than an established result.
