\section{Conclusion}

This survey has argued that predictive world models from video should be understood by the role their predictions play, rather than by visual generation quality alone. Future-frame prediction is an important starting point, especially when it is action-conditioned or uncertainty-aware, but it becomes world-model relevant when it supports reasoning about interventions, physical change, planning, controllability, or evaluation beyond reconstruction. The central perspective of the survey is therefore concept-driven: modeling goal, representation, dynamics, conditioning, and evaluation jointly determine what kind of world model a system learns.

Across the reviewed literature, the field shows a progression from pixel-level video prediction toward increasingly structured and interactive forms of generative simulation. Object-centric and relational models introduce entities, interactions, and hidden physical variables as useful abstractions for physical dynamics. Latent, token, and feature-space world models shift prediction into compact or learned states that can support planning, policy learning, imagined experience, or behavior-oriented evaluation. Diffusion-based and interactive models revisit image-space simulation and controllable video generation, while also making the abstraction-fidelity trade-off more explicit. Evaluation then becomes a cross-cutting concern: visual fidelity, control return, physical consistency, causal validity, and controllability test different aspects of world-model behavior.

The main unresolved challenge is to build models that are simultaneously visually realistic, dynamically predictive, physically consistent, and interactively controllable. The preceding sections suggest that progress along one axis does not guarantee progress along the others: realistic videos may remain physically inconsistent, compact latent states may omit control-relevant details, and controllable rollouts may still lack grounded physical semantics. A cautious path forward is to treat video world modeling as both a representation-learning problem and an evaluation problem.

This view leads to several design principles for future work. A world model should specify what its state is, whether pixels, objects, latent variables, tokens, or feature embeddings. It should specify which interventions it supports, such as real actions, robot commands, language instructions, latent controls, or counterfactual changes. Visual fidelity is insufficient without rollout validity, because plausible frames can still fail to preserve state identity, physical constraints, or action consequences. Control success is also insufficient without physical and causal robustness, because reward-defined tasks can hide simulator errors outside the evaluated policy distribution. Future benchmarks should therefore jointly test prediction, control, long-horizon consistency, and physical-law generalization rather than treating any single metric as evidence of world-model validity.
